Composite Bayesian inference

### 2019-02-08 papers/composite_inference/draft.tex
Init Madrugada

\newcommand{\E}{\mathbb{E}}
\newcommand{\lda}{{\boldsymbol{\lambda}}}
%\newcommand{\matphi}{\boldsymbol{\Phi}} 
%\def\x{{\mathbf{x}}}
%\def\u{{\mathbf{u}}}
%\def\p{{\bar{\mathbf{p}}}}
%\def\q{{\bar{\mathbf{q}}}}

@@

% p(x,y|theta) = p(x|y,theta)p(y|theta)

@@

Let $Y$ an observable multivariate random variable with sampling distribution $p(y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is assumed to be a finite set for simplicity. Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $x$:
$$
L(x) = p(y|x)
.
$$

This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated. A natural workaround known as data reduction is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
$$
\ell(x) = p(z|x)
.
$$

Substituting $L(x)$ with $\ell(x)$ boils down to restricting the feature space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information. While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction. To make the trick statistically more efficient, we may extract several features, $z_i=f_i(y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = p(z_i|x)$ that they elicit.

If we see the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. One that is appealing as it tends to produce single peaked distributions is the {\em generalized logarithmic opinion pool}:
\begin{equation}
\label{eq:log_pool}
p_\lda(x|y) \propto \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
\end{equation} 
where $\pi(x)$ is a prior distribution, and $\lda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lda(x_1|y)\leq p_\lda(x_2|y)$. Also note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.

%This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are upper bounded. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|y)\leq p_\lambda(x_2|y)$. Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.

@@

The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding  the form: 
$$
p_\lda(x|y)\propto \pi(x) L^c_\lda(x),
$$
where the distribution $\pi(x)$ plays the role of a prior, and the function:
\begin{equation}
\label{eq:comp_lik}
L^c_\lda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
\end{equation} 
plays that of a likelihood function. The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}. The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{appendix:conditional}. The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.

CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lda\equiv 1$. While the above derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method. However, unitary CL weights do not guarantee optimal prediction performance.

@@

A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$y$. Given two putative target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ is computed by simply adding up single-feature contributions: 
$$
\log \frac{p_\lda(x_1|y)}{p_\lda(x_0|y)}
= 
\underbrace{\log \frac{\pi(x_1)}{\pi(x_0)}}_{\text{offset}}
+ \sum_i \underbrace{\lambda_i \log \frac{p(z_i|x_1)}{p(z_i|x_0)}}_{\text{feature contribution}},
$$
therefore it is straightforward to assess which features contribute most to a particular hypothesis comparison (either positively or negatively), see possible disagreements between features, and corroborate predictive probabilities with feature-level testable information.

Each feature contributes to an hypothesis comparison proportionally to its associated log-likelihood ratio test statistic. Contrary to statistical hypothesis testing, likelihood ratios are not converted into $p$-values but are directly combined into a predictive probability ratio. Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction ($\lda\equiv 1$), and may therefore account for mutual feature dependence if tuned appropriately. Keeping the analogy with hypothesis testing, composite weights may be thought of as the counterpart of multiple testing correction \cite{Benjamini-10}.

@@

%So far, we have assumed that the composite weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) are known.

Let us now turn to the problem of tuning the composite weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) as part of the method calibration. It is assumed at this stage that the feature generative models are known, which may require a pre-training stage that should not cause technical difficulties as it may use standard parameter estimation techniques.

@@

Given a labeled training dataset ${\cal D}=\{(x_k, y_k), k=1,\ldots,N\}$, the most direct approach to tune the composite weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'') under the composite predictive distribution:
\begin{equation}
\label{eq:train_likelihood}
\max_{\lda\succeq 0} U(\lda)
\equiv\sum_{k=1}^N \log p_\lda(x_k|y_k).
\end{equation}

From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\lda(x|y)$ spanned by the weights~$\lda$ is the exponential family with natural parameter~$\lda$ and basis functions given by the feature-based log-likelihoods:
$$
p_\lda(x|y) = \pi(x) \exp[\lda^\top \mathbf{L}(x,y) - A(\lda,y)],
$$
with:
$$
L_i(x,y) \equiv \log p(z_i|x),
\qquad
A(\lda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\lda^\top \mathbf{F}(x,y)}.
$$

It follows that the utility function $U(\lda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:maxent} for details. It also implies that maximizing $U(\lda)$ is dual to a minimum KL~divergence problem \cite{Csiszar-84}:
$$
\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,y) \log \frac{p(x,y)}{\pi(x)h(y)},
$$
subject to the constraints $p(y)=h(y)$ and $\E_p[\mathbf{L}] \succeq \E_h[\mathbf{L}]$, where 
${\cal Y}=\{y_1,\ldots,y_N\}$ is the set of training data values, ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$, $h(x,y)$ is the joint empirical distribution of targets and data, and $h(y)$ is the corresponding marginal distribution.

Therefore, the optimal composite model may be interpreted as the $I$-projection of the independence distribution $\pi(x)h(y)$ onto a set of admissible joint distributions defined by two types of contraints. The marginal constraint $p(y)=h(y)$, on the one hand,  ensures that only the conditional distribution $p(x|y)$ is ``updated'', making the $I$-projection a special case of maximum conditional entropy \cite{BergerA-96}. The mean log-likelihood value constraints $\E_p[\mathbf{L}] \succeq \E_h[\mathbf{L}]$, on the other hand, encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature. This is weaker a constraint than assuming that the features are distributed according to the generative models $p(z_i|x)$: it only assumes that the ``description length'' \cite{Grunwald-07} achieved by each agent is truthworthy.

@@

As is customary and is generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$ (obtaining balanced classes if the prior is uniform). This does not affect the above maths provided that the data weights are reflected in the empirical distribution~$h(x,y)$.

Other variants of the training scheme that we mention here but do not particularly recommend for interpretability include adding basis functions to the exponential family. For instance, class-dependent offsets may be learned concurrently with the composite weights, which amounts to learning the ``prior'' in the training phase. It is also possible to replace the mean log-likelihood inequality constraints with equality constraints: the maximization of $U(\lda)$ is then unconstrained, therefore some trained composite weights may be negative. Finally, the mean log-likelihood constraints may be expressed for each class rather than being averaged wrt the prior: this leads to an extension that we call ``super composite likelihood'', in which weights become class-dependent (see Appendix~\ref{appendix:super}).

@@

%We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(y)p_\lambda(x|y)dy = \pi(x)$. This ``deforms'' the prior to improve the training score. But why is it important to enforce this constraint? Again, it all makes things just less interpretable.

@@

%Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$. We expect to have $\lambda=1$ but it's not necessarily the case. It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active). This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold. So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.

@@

As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
\begin{equation}
\label{eq:cond_feat_lik}
\ell_i(x) = p(z_i|x,z^c_i).
\end{equation}

@@

Let $\pi(x)$ some prior distribution on~$X$. Optimal prediction aims to select the joint distribution $p(x,y)$ that minimizes the KL divergence $D(p(x,y)\|\pi(x)p(y))$:
$$
D(p) = -\int p(x,y) \log \frac{p(x|y)}{\pi(x)} dx dy,
$$ subject to feature mean-value constraints of the form:
$$
\int p(x,y) f(x,y) dx dy = \mu.
$$

@@

One important special case of this problem arises when imposing the additional constraint that $p(y)=h(y)$ is known. This is the {\em maximum conditional entropy} method \cite{BergerA-96} which we review hereafter.

@@

$$
D(p) = -\int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
$$

Lagrangian:
$$
{\cal L}(p,\lambda) = D(p) + \lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - \mu)
$$

Optimal $p$ at fixed $\lambda$:
$$
\frac{\partial{\cal L}}{\partial p}
=
- h(y) \log \frac{p(x|y)}{\pi(x)} + h(y) \lambda^\top f(x,y) + (\kappa(y)-1)h(y)
$$
$$
\Rightarrow
\quad
p_{\lambda}(x|y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
$$
with:
$$
z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
$$

Dual:
$$
\psi(\lambda) = \lambda^\top \mu - \int h(y) \log z(\lambda, y) dy 
$$

@@

$$
\nabla\psi
= \mu - \int h(y) \frac{\nabla z}{z} dy
$$

$$
\nabla\nabla^\top\psi
= - \int h(y) \left(
\frac{\nabla \nabla^\top z}{z} 
- \frac{\nabla z \nabla z^\top}{z^2}
\right)
dy
$$

Note the equivalent expression of the dual function as:
$$
\psi(\lambda)
=
\int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
= D(h\|\pi(x)h(y)) - D(h\|p_\lambda)
$$

@@

Let $m_F(y)$ and $V_F(y)$ respectively denote the mean and variance of $f(x,y)$ conditional on~$y$. Denoting by $\langle \rangle$ the empirical mean wrt $y$:
$$
\nabla \psi = \mu - \langle m_F \rangle
$$

@@

In general, we have more features than classes for good prediction performance so each matrix $V_F(y)$ is singular and tends to be smaller (in terms of spectral radius) as the number of features increases. The Hessian is a sample average of these matrices and will thus follow the same trend.

@@

Let us consider a reference target value $x_0$. By definition,
$$
p_\lambda(x_0|y) = \frac{\pi(x_0)}{z(\lambda,y)} e^{\lambda^\top f(x_0,y)}
$$

Therefore,
$$
\log z(\lambda,y) = \lambda^\top f(x_0,y) - \log \frac{p_\lambda(x_0|y)}{\pi(x_0)}
$$

We can use this fact to rewrite the dual function:
$$
\psi(\lambda) = \lambda^\top (\mu - \mu_0) + \int h(y) \log \frac{p_\lambda(x_0|y)}{\pi(x_0)} dy
$$
with: 
$$
\mu_0 = \int h(y) f(x_0, y) dy
$$

The first term $\lambda^\top (\mu-\mu_0)$ is essentially the same thing as the utility considered in my previous version of composite likelihood training (which was maximized on the simplex). Utility is label-dependent as it involves the moments. If the basis consists of feature-based log-likelihood functions, it measures the discrepancy vs the reference and pushes towards large coefficient values. Cross-entropy appears here as a regularized version of utility. The regularization term $E[\log p_\lambda(x_0|y)$ replaces the hard constraint that $\lambda$ lie in the simplex: it forces the reference class to be probable for any input, so it will tend to discard predictions that are peaked on a single class if it's not $x_0$. This term acts as a callback force.

@@

In this case, the constraint $p(y)=h(y)$ no longer applies. Consider:
$$
D(p,m) = \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dx dy
$$
and note that the prediction objective is: $D(p) = \min_m D(p,m)$. This is the Blahut-Arimoto trick, which enables us to consider an alternate optimization strategy wrt $p$ and $m$, respectively.

At fixed $p$, this is a maximum (joint) relative entropy problem. Lagrangian...
$$
{\cal L}(p,m,\lambda)
=
- \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dxdy
+
\lambda^\top \left( 
\int p(x,y) f(x,y) dydy - \mu 
\right) \\
$$

The derivative is given by:
$$
\frac{\partial\cal L}{\partial p(x,y)}
= 
- \log \frac{p(x,y)}{\pi(x)m(y)} - 1
+ \lambda^\top f(x,y)
$$
hence the optimal model at fixed $m$ has the form:
$$
p_{\lambda,m}(x,y) = \frac{1}{z(\lambda,m)} \pi(x) m(y) e^{\lambda^\top f(x,y)} 
$$
with:
$$
z(\lambda,m) = \int \pi(x) m(y) e^{\lambda^\top f(x,y)} dx dy
$$

Note that the induced posterior distribution $p_{\lambda,m}(x|y)\propto \pi(x)p_{\lambda,m}(y|x)$ has the same form as in the conditional maxent case (if using the very same moment constraints).

@@

$$
\nabla z = \int \pi(x) m(y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
$$

$$
\nabla\nabla^\top z = \int \pi(x)m(y) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dxdy
$$

@@

Equivalently, for any distribution $h(x,y)$ that satisfies the moment constraint, we have:
$$
\psi(\lambda,m) 
=
\int h(x,y) \log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(y)} dxdy
= D(h\|\pi_xm_y) - D(h\|p_{\lambda,m})
$$
showing that this is a generative learning problem!

The BA algorithm alternates between estimating $\lambda$ at fixed $m$ via the maximization of $\psi(\lambda,m)$ and estimating $m$ at fixed $\lambda$ by marginalization:
$$
m(y) = \int p_{\lambda,m}(x,y) dx
$$

>>>>

\def\x{{\mathbf{x}}}
\def\y{{\mathbf{y}}}
\def\y{{\mathbf{z}}}
\newcommand{\blambda}{{\boldsymbol{\lambda}}}
\newcommand{\bell}{{\boldsymbol{\ell}}}
\newcommand{\E}{\mathbb{E}}
%\newcommand{\matphi}{\boldsymbol{\Phi}} 
%\def\p{{\bar{\mathbf{p}}}}
%\def\q{{\bar{\mathbf{q}}}}

@@

@@

Let $\mathbf{Y}$ an observable multivariate random variable with sampling distribution $p(\y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is assumed to be a finite set for simplicity. Given an experimental outcome $\y$, the likelihood is the sampling distribution evaluated at $\y$, seen as a function of $x$:
$$
L(x) = p(\y|x)
.
$$

This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated. A natural workaround known as data reduction is to extract some lower-dimensional representation $z=f(\y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
$$
\ell(x) = p(z|x)
.
$$

Substituting $L(x)$ with $\ell(x)$ boils down to restricting the feature space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information. While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction. To make the trick statistically more efficient, we may extract several features, $z_i=f_i(\y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = p(z_i|x)$ that they elicit.

If we see the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. One that is appealing as it tends to produce single peaked distributions is the {\em generalized logarithmic opinion pool}:
\begin{equation}
\label{eq:log_pool}
p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
\end{equation} 
where $\pi(x)$ is a prior distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$. Also note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.

%This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are upper bounded. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|\y)\leq p_\lambda(x_2|\y)$. Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.

@@

The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding  the form: 
$$
p_\blambda(x|\y)\propto \pi(x) L^c_\blambda(x),
$$
where the distribution $\pi(x)$ plays the role of a prior, and the function:
\begin{equation}
\label{eq:comp_lik}
L^c_\blambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
\end{equation} 
plays that of a likelihood function. The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}. The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{appendix:conditional}. The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.

CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$. While the above derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method. However, unitary CL weights do not guarantee optimal prediction performance.

@@

A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$y$. Given two putative target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ is computed by simply adding up single-feature contributions: 
$$
\log \frac{p_\blambda(x_1|\y)}{p_\blambda(x_0|\y)}
= 
\underbrace{\log \frac{\pi(x_1)}{\pi(x_0)}}_{\text{offset}}
+ \sum_i \underbrace{\lambda_i \log \frac{p(z_i|x_1)}{p(z_i|x_0)}}_{\text{feature contribution}},
$$
therefore it is straightforward to assess which features contribute most to a particular hypothesis comparison (either positively or negatively), see possible disagreements between features, and corroborate predictive probabilities with feature-level testable information.

Each feature contributes to an hypothesis comparison proportionally to its associated log-likelihood ratio test statistic. Contrary to statistical hypothesis testing, likelihood ratios are not converted into $p$-values but are directly combined into a predictive probability ratio. Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction ($\blambda\equiv 1$), and may therefore account for mutual feature dependence if tuned appropriately. Keeping the analogy with hypothesis testing, composite weights may be thought of as the counterpart of multiple testing correction \cite{Benjamini-10}.

@@

%So far, we have assumed that the composite weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) are known.

Let us now turn to the problem of tuning the composite weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) as part of the method calibration. It is assumed at this stage that the feature generative models are known, which may require a pre-training stage that should not cause technical difficulties as it may use standard parameter estimation techniques.

@@

Given a labeled training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct approach to tune the composite weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'') under the composite predictive distribution:
\begin{equation}
\label{eq:train_likelihood}
\max_{\blambda\succeq 0} U(\blambda)
\equiv\sum_{k=1}^N \log p_\blambda(x_k|\y_k).
\end{equation}

From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\blambda(x|\y)$ spanned by the weights~$\blambda$ is the exponential family with natural parameter~$\blambda$ and basis functions given by the feature-based log-likelihoods:
$$
p_\blambda(x|\y) = \pi(x) \exp[\blambda^\top \bell(x,\y) - A(\blambda,y)],
$$
with:
$$
\ell_i(x,\y) \equiv \log p(z_i|x),
\qquad
A(\blambda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
$$

It follows that the utility function $U(\blambda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:maxent} for details. It also implies that maximizing $U(\blambda)$ is dual to a minimum KL~divergence problem \cite{Csiszar-84}:
$$
\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{\pi(x)h(\y)},
$$
subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where 
${\cal Y}=\{\y_1,\ldots,\y_N\}$ is the set of training data values, ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$, $h(x,\y)$ is the joint empirical distribution of targets and data, and $h(\y)$ is the corresponding marginal distribution.

Therefore, the optimal composite model may be interpreted as the $I$-projection of the independence distribution $\pi(x)h(\y)$ onto a set of admissible joint distributions defined by two types of contraints. The marginal constraint $p(\y)=h(\y)$, on the one hand,  ensures that only the conditional distribution $p(x|\y)$ is ``updated'', making the $I$-projection a special case of maximum conditional entropy \cite{BergerA-96}. The mean log-likelihood value constraints $\E_p[\bell] \succeq \E_h[\bell]$, on the other hand, encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature. This is weaker a constraint than assuming that the features are distributed according to the generative models $p(z_i|x)$: it only assumes that the ``description length'' \cite{Grunwald-07} achieved by each agent is truthworthy.

@@

As is customary and is generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$ (obtaining balanced classes if the prior is uniform). This does not affect the above maths provided that the data weights are reflected in the empirical distribution~$h(x,\y)$.

Other variants of the training scheme that we mention here but do not particularly recommend for interpretability include adding basis functions to the exponential family. For instance, class-dependent offsets may be learned concurrently with the composite weights, which amounts to learning the ``prior'' in the training phase. It is also possible to replace the mean log-likelihood inequality constraints with equality constraints: the maximization of $U(\blambda)$ is then unconstrained, therefore some trained composite weights may be negative. Finally, the mean log-likelihood constraints may be expressed for each class rather than being averaged wrt the prior: this leads to an extension that we call ``super composite likelihood'', in which weights become class-dependent (see Appendix~\ref{appendix:super}).

@@

%We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(\y)p_\lambda(x|y)dy = \pi(x)$. This ``deforms'' the prior to improve the training score. But why is it important to enforce this constraint? Again, it all makes things just less interpretable.

@@

%Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(\y|x)^\lambda$. We expect to have $\lambda=1$ but it's not necessarily the case. It is if $h(\y)=\int\pi(x)p(\y|x)dx$ because then $h(\y)p(x|y)=\pi(x)p(\y|x)$ so the constraint is verified (and active). This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold. So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.

@@

As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(\y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(\y)$, yielding the more general form:
\begin{equation}
\label{eq:cond_feat_lik}
\ell_i(x) = p(z_i|x,z^c_i).
\end{equation}

@@

Let $\pi(x)$ some prior distribution on~$X$. Optimal prediction aims to select the joint distribution $p(x,\y)$ that minimizes the KL divergence $D(p(x,\y)\|\pi(x)p(\y))$:
$$
D(p) = -\int p(x,\y) \log \frac{p(x|\y)}{\pi(x)} dx d\y,
$$ subject to feature mean-value constraints of the form:
$$
\int p(x,\y) f(x,\y) dx d\y = \mu.
$$

@@

One important special case of this problem arises when imposing the additional constraint that $p(\y)=h(\y)$ is known. This is the {\em maximum conditional entropy} method \cite{BergerA-96} which we review hereafter.

@@

$$
D(p) = -\int h(\y)p(x|\y) \log \frac{p(x|\y)}{\pi(x)} dxdy
$$

Lagrangian:
$$
{\cal L}(p,\lambda) = D(p) + \lambda^\top (\int h(\y)p(x|\y)f(x,y)dxdy - \mu)
$$

Optimal $p$ at fixed $\lambda$:
$$
\frac{\partial{\cal L}}{\partial p}
=
- h(\y) \log \frac{p(x|\y)}{\pi(x)} + h(\y) \lambda^\top f(x,y) + (\kappa(\y)-1)h(\y)
$$
$$
\Rightarrow
\quad
p_{\lambda}(x|\y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
$$
with:
$$
z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
$$

Dual:
$$
\psi(\lambda) = \lambda^\top \mu - \int h(\y) \log z(\lambda, y) dy 
$$

@@

$$
\nabla\psi
= \mu - \int h(\y) \frac{\nabla z}{z} dy
$$

$$
\nabla\nabla^\top\psi
= - \int h(\y) \left(
\frac{\nabla \nabla^\top z}{z} 
- \frac{\nabla z \nabla z^\top}{z^2}
\right)
dy
$$

Note the equivalent expression of the dual function as:
$$
\psi(\lambda)
=
\int h(x,y) \log \frac{p_\lambda(x|\y)}{\pi(x)} dx dy
= D(h\|\pi(x)h(\y)) - D(h\|p_\lambda)
$$

@@

Let $m_F(\y)$ and $V_F(\y)$ respectively denote the mean and variance of $f(x,y)$ conditional on~$y$. Denoting by $\langle \rangle$ the empirical mean wrt $y$:
$$
\nabla \psi = \mu - \langle m_F \rangle
$$

@@

In general, we have more features than classes for good prediction performance so each matrix $V_F(\y)$ is singular and tends to be smaller (in terms of spectral radius) as the number of features increases. The Hessian is a sample average of these matrices and will thus follow the same trend.

@@

Let us consider a reference target value $x_0$. By definition,
$$
p_\lambda(x_0|\y) = \frac{\pi(x_0)}{z(\lambda,y)} e^{\lambda^\top f(x_0,y)}
$$

Therefore,
$$
\log z(\lambda,y) = \lambda^\top f(x_0,y) - \log \frac{p_\lambda(x_0|\y)}{\pi(x_0)}
$$

We can use this fact to rewrite the dual function:
$$
\psi(\lambda) = \lambda^\top (\mu - \mu_0) + \int h(\y) \log \frac{p_\lambda(x_0|\y)}{\pi(x_0)} dy
$$
with: 
$$
\mu_0 = \int h(\y) f(x_0, y) dy
$$

The first term $\lambda^\top (\mu-\mu_0)$ is essentially the same thing as the utility considered in my previous version of composite likelihood training (which was maximized on the simplex). Utility is label-dependent as it involves the moments. If the basis consists of feature-based log-likelihood functions, it measures the discrepancy vs the reference and pushes towards large coefficient values. Cross-entropy appears here as a regularized version of utility. The regularization term $E[\log p_\lambda(x_0|\y)$ replaces the hard constraint that $\lambda$ lie in the simplex: it forces the reference class to be probable for any input, so it will tend to discard predictions that are peaked on a single class if it's not $x_0$. This term acts as a callback force.

@@

In this case, the constraint $p(\y)=h(\y)$ no longer applies. Consider:
$$
D(p,m) = \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(\y)} dx dy
$$
and note that the prediction objective is: $D(p) = \min_m D(p,m)$. This is the Blahut-Arimoto trick, which enables us to consider an alternate optimization strategy wrt $p$ and $m$, respectively.

At fixed $p$, this is a maximum (joint) relative entropy problem. Lagrangian...
$$
{\cal L}(p,m,\lambda)
=
- \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(\y)} dxdy
+
\lambda^\top \left( 
\int p(x,y) f(x,y) dydy - \mu 
\right) \\
$$

The derivative is given by:
$$
\frac{\partial\cal L}{\partial p(x,y)}
= 
- \log \frac{p(x,y)}{\pi(x)m(\y)} - 1
+ \lambda^\top f(x,y)
$$
hence the optimal model at fixed $m$ has the form:
$$
p_{\lambda,m}(x,y) = \frac{1}{z(\lambda,m)} \pi(x) m(\y) e^{\lambda^\top f(x,y)} 
$$
with:
$$
z(\lambda,m) = \int \pi(x) m(\y) e^{\lambda^\top f(x,y)} dx dy
$$

Note that the induced posterior distribution $p_{\lambda,m}(x|\y)\propto \pi(x)p_{\lambda,m}(\y|x)$ has the same form as in the conditional maxent case (if using the very same moment constraints).

@@

$$
\nabla z = \int \pi(x) m(\y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
$$

$$
\nabla\nabla^\top z = \int \pi(x)m(\y) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dxdy
$$

@@

Equivalently, for any distribution $h(x,y)$ that satisfies the moment constraint, we have:
$$
\psi(\lambda,m) 
=
\int h(x,y) \log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(\y)} dxdy
= D(h\|\pi_xm_y) - D(h\|p_{\lambda,m})
$$
showing that this is a generative learning problem!

The BA algorithm alternates between estimating $\lambda$ at fixed $m$ via the maximization of $\psi(\lambda,m)$ and estimating $m$ at fixed $\lambda$ by marginalization:
$$
m(\y) = \int p_{\lambda,m}(x,y) dx
$$

The MIL seems intractable if we don't impose constraints on $m(\y)$. It becomes feasible by restricting $m(\y)$ to have a support limited to the training data, similarly to the conditional maxent implementation:
$$
m(\y) = \sum_k w_k \delta(\y-\y_k),
$$
hence $m$ is determined by a finite set of weights $w_k$. In practice, the optimal weights look very sparse.
